Multiple imputation of incomplete multilevel data using Heckman selection models

### 2022-12-16 1.Paper/Paper.Rmd
manuscript update

Under the BVN assumption, the parameters of the Heckman model coefficients can be estimated using the two-step Heckman method (H2S) [@heckman1976] or the full information maximum likelihood method (FIML).[@amemiya1984] However, both methods may lead to inconsistent estimators when the true underlying distribution is not the BVN. [@gomes_etal20] To overcome this problem, other approaches have been proposed that relax the distribution assumptions, among them copula models. [@smith03] The copula approach uses a function, known as a copula, that joins the marginal distribution of the error terms of the selection and outcome equation which are specified separately. The Heckman model can be expressed with the following bivariate normal copula:
$$ F(\boldsymbol{r^*_{i}},\boldsymbol{y^*_{i}}) =\Phi\left(\boldsymbol{r^*_{i}}-\boldsymbol{X^S_{i}}\boldsymbol{\beta^S_{i}},\frac{\boldsymbol{y^*_{i}}-\boldsymbol{X^O_{i}}\boldsymbol{\beta^O_{i}}}{\sigma_i}, \rho_i\right),$$
where $\Phi$ is the bivariate standard normal cumulative distribution function.
Thus, to estimate the parameters of the Heckman method, it is sufficient to specify the marginal distributions of the error terms, and link them with a suitable copula function. In our imputation method we estimate the Heckman model using the copula method, as in real life data generating mechanism may deviate a BVN distribution.

@@

In this step, we pool the parameters $\boldsymbol{\widehat{\theta_i}}$ with a random effects meta-analysis model using only the clusters with observable information, i.e., those with no systematically missing outcome. In particular, we pool the $p$ coefficients of the $\boldsymbol{\beta^O}$ in the outcome equation and estimate a multivariate random effects meta-analysis model with them. Similarly we combine all $q$ coefficient parameters of the $\boldsymbol{\beta^S}$ in the selection equation. 
Thus, we can denote the study specific coefficients
$$\boldsymbol{\hat{\beta^O}_i} = \boldsymbol{\beta^O} +\boldsymbol{ b^O_i} + \boldsymbol{\epsilon'^O_{i}}$$
$$\boldsymbol{\hat{\beta^S}_i} = \boldsymbol{\beta^S} + b^S_i + \boldsymbol{\epsilon'^S_{i}}$$
using the random effects 
 $b^O_i \sim N(0,\boldsymbol{\Psi^O})$ and $b^S_i\sim N(0,\boldsymbol{\Psi^S})$ with $\epsilon'^O_{i}\sim N(0,\sigma^O_{i}(X^O_i{X^O_i}^T))$ and $\epsilon'^S_{i}\sim N(0,\sigma^S_{i}(X^S_i{X^S_i}^T))$. From the meta model we obtain the estimates of parameters $\boldsymbol{\beta^S}$,$\boldsymbol{\beta^O}$, $\boldsymbol{\Psi^S}$,$\boldsymbol{\Psi^O}$.

Assuming that  $\sigma_i$ and $\rho_i$ are random variables comming from a populational distribution, they can express as:
$$ log(\sigma_i) \sim N( log(\sigma),\psi^{\sigma})$$
$$ tanh^{-1}(\rho_i)\sim N( tanh^{-1}(\rho),\psi^{\rho})$$ 
For each of them, we perform a univariate random effects meta_analysis. The model for $\sigma_i$ is given by the model:
$$log(\hat\sigma_i) = \sigma + b^{\sigma}_i + \epsilon^{\sigma}_{i}$$
with $b_i^{\sigma} \sim N(0,\psi^{\sigma})$, and $\epsilon_i^{\sigma} \sim N(0,var(log(\hat{\sigma}_i)))$ and  the model for $\rho_i$ is:
$$tanh^{-1}(\hat\rho_i) = \rho + b^{\rho}_i + \epsilon^{\rho}_{i}$$
with $b_i^{\rho} \sim N(0,\psi^{\rho})$, and $\epsilon_i^{\rho} \sim N(0,var(tanh^{-1}(\hat{\rho}_i)))$.

@@

From the meta-analysis model, we obtain the marginal estimates $\boldsymbol{\widehat{\Theta}}=\{\widehat{\boldsymbol{\beta^S}},\widehat{\boldsymbol{\beta^O}},\hat{\sigma},\hat{\rho}\}$ and the between-cluster variance matrix $\boldsymbol{\widehat{\psi}}$,i.e. variance-covariance matrix of the random effects, with their corresponding variance-covariance matrices $\boldsymbol{\widehat{S_{\Theta}}}$ and $\boldsymbol{\widehat{S_{\psi}}}$, which are used to draw the $\boldsymbol{\Theta^*}$ and $\boldsymbol{\psi^*}$ parameters, from their posterior distribution as follows: [@resche-rigon2013]

@@

We generated the data from a Heckman selection model with bivariate normal distribution error terms. For simplicity we started from "basic scenario" where the database collected information from $N=10$ clusters of $n_i=1000$  individuals. However, we altered both $N$ and $n_i$ in sensitivity analyses (Table \ref{tab:scenarios}). 
```{r scenarios,  echo = FALSE, warning = FALSE,cache = TRUE}
data2 <- read.table(stringsAsFactors = FALSE, header = TRUE, sep="/", text =
'Scenario/	Incomplete variable/	Rho/	N; $n_i$/	Missing process
Base/	Continuous/	0.6/	10;1000/	Heckman (BVN)
M(N)AR/Continuous, Binary/ 0 (MAR), 0.3,0.6,0.9 (MNAR)/	10;1000/Heckman (BVN)
Size and cluster number/	Continuous/	0.6/	10;50,   10;100,   10;1000,  50;1000,  100;1000/	Heckman (BVN)
Distribution deviations/	Continuous/	0.6/	10;1000	/Heckman (BVN), Heckman (t-skew), Explicit'
)
options(kableExtra.latex.load_packages = FALSE) 
library(kableExtra)
kbl(data2, booktabs = T, align="l",
    caption = "Data generation scenarios",
    col.names = c("Scenario","Incomplete variable", "$\\rho$", "$N;n_i$","Missing process"),
    row.names=FALSE,escape = FALSE)%>%
    column_spec(c(1), width = "2,5cm")%>%
    column_spec(c(2,4), width = "1,5cm")%>%
    column_spec(c(3), width = "2,9cm")%>%
    column_spec(c(5), width = "2.7cm")%>%
  kable_styling(latex_options = c("striped"))
```
For each dataset, we generated $X_{1i}$ a treatment indicator variable from a Bernoulli distribution with a probability of treatment on each cluster equal to 0.6. Next, we simulated the mean of two continuous covariates from a multivariate normal distribution $\mu_h\sim N(\begin{psmallmatrix}0\\0\end{psmallmatrix},\begin{psmallmatrix}0.2 & 0.015\\0.015 & 0.2\end{psmallmatrix})$, with $h=\{2,3\}$. We then simulated for each cluster a baseline covariate $X_{2i}\sim N(\mu_2,1)$ and a exclusion restriction variable  $X_{3i} \sim N(\mu_3,0.5)$.

@@

As regards the error terms, they were bivariate normal distributed as:
\[\begin{pmatrix}
\epsilon^O_{i} \\
\epsilon^S_{i}
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\begin{pmatrix}
\sigma^2_i & \rho\sigma_i \\
\rho\sigma_i & 1
\end{pmatrix}\right)
\]
whose $\sigma_i^2$ is variable across clusters and distributed as $log(\sigma_i) \sim N(0,0.05)$.

>>>>

Under the BVN assumption, the parameters of the Heckman model coefficients can be estimated using the two-step Heckman method (H2S) [@heckman1976] or the full information maximum likelihood method (FIML).[@amemiya1984] However, both methods may lead to inconsistent estimators when the true underlying distribution is not the BVN. [@gomes_etal20] To overcome this problem, other approaches have been proposed that relax the distribution assumptions, among them copula models. [@smith03] The copula approach uses a function, known as a copula, that joins the marginal distribution of the error terms of the selection and outcome equation which are specified separately. The Heckman model can be expressed with the following bivariate normal copula:
$$ F(\boldsymbol{r^*_{i}},\boldsymbol{y^*_{i}}) =\Phi\left(\boldsymbol{r^*_{i}}-\boldsymbol{X^S_{i}}\boldsymbol{\beta^S_{i}},\frac{\boldsymbol{y^*_{i}}-\boldsymbol{X^O_{i}}\boldsymbol{\beta^O_{i}}}{\sigma_i}, \rho_i\right),$$
where $\Phi$ is the bivariate standard normal cumulative distribution function.
Thus, to estimate the parameters of the Heckman method, it is sufficient to specify the marginal distributions of the error terms, and link them with a suitable copula function. In our imputation method we estimate the Heckman model using the copula method for non-random selection [@wojtys2018] which is a flexible approach that allows the specification of different parametric distributions of the selection and outcome variables and also different types of dependence structure between the two equations.

@@

In this step, we pool the parameters $\boldsymbol{\widehat{\theta_i}}$ with a random effects meta-analysis model using only the clusters with observable information, i.e., those with no systematically missing outcome. In particular, we pool the $p$ coefficients of the $\boldsymbol{\beta^O}$ in the outcome equation and estimate a multivariate random effects meta-analysis model with them. Similarly we combine all $q$ coefficient parameters of the $\boldsymbol{\beta^S}$ in the selection equation. 
Thus, we can denote the study specific coefficients
$$\boldsymbol{\hat{\beta^O}_i} = \boldsymbol{\beta^O} +\boldsymbol{ b^O_i} + \boldsymbol{\epsilon'^O_{i}}$$
$$\boldsymbol{\hat{\beta^S}_i} = \boldsymbol{\beta^S} + b^S_i + \boldsymbol{\epsilon'^S_{i}}$$
using the random effects 
 $b^O_i \sim N(0,\boldsymbol{\Psi^O})$ and $b^S_i\sim N(0,\boldsymbol{\Psi^S})$ with $\epsilon'^O_{i}\sim N(0,\sigma^O_{i}(X^O_i{X^O_i}^T))$ and $\epsilon'^S_{i}\sim N(0,\sigma^S_{i}(X^S_i{X^S_i}^T)^{-1})$.

We assume that  $\sigma_i$ and $\rho_i$ are random variables coming from a population distribution, so we can express them as:
$$ log(\sigma_i) \sim N( log(\sigma),\psi^{\sigma})$$
$$ tanh^{-1}(\rho_i)\sim N( tanh^{-1}(\rho),\psi^{\rho})$$ 
For each of them, we perform a univariate random effects meta-analysis. The model for $\sigma_i$ is given by the model:
$$log(\hat\sigma_i) = \sigma + b^{\sigma}_i + \epsilon^{\sigma}_{i}$$
with $b_i^{\sigma} \sim N(0,\psi^{\sigma})$, and $\epsilon_i^{\sigma} \sim N(0,var(log(\hat{\sigma}_i)))$ and  the model for $\rho_i$ is:
$$tanh^{-1}(\hat\rho_i) = \rho + b^{\rho}_i + \epsilon^{\rho}_{i}$$
with $b_i^{\rho} \sim N(0,\psi^{\rho})$, and $\epsilon_i^{\rho} \sim N(0,var(tanh^{-1}(\hat{\rho}_i)))$.

In the case of an incomplete binary variable, the $\sigma$ parameter is not specified, so it is not necessary to perform a random effects meta-analysis model of $\sigma$ or to include it in any of the following steps in the imputation model.

@@

From the meta-analysis model, we obtain the marginal estimates $\boldsymbol{\widehat{\Theta}}=\{\widehat{\boldsymbol{\beta^S}},\widehat{\boldsymbol{\beta^O}},\hat{\sigma},\hat{\rho}\}$ and the between-cluster variance matrix $\boldsymbol{\widehat{\psi}}=\{\boldsymbol{\Psi^S},\boldsymbol{\Psi^O},\psi^{\sigma},\psi^{\rho}\}$, i.e. variance-covariance matrix of the random effects, with their corresponding variance-covariance matrices $\boldsymbol{\widehat{S_{\Theta}}}$ and $\boldsymbol{\widehat{S_{\psi}}}$, which are used to draw the $\boldsymbol{\Theta^*}$ and $\boldsymbol{\psi^*}$ parameters, from their posterior distribution as follows: [@resche-rigon2013]

@@

We generated the data from a Heckman selection model with bivariate normal distribution error terms. For simplicity we started from "basic scenario" where the database collected information from $N=10$ clusters of $n_i=1000$  individuals. However, we altered both $N$ and $n_i$ in sensitivity analyses (Table \ref{tab:scenarios}). 
```{r scenarios,  echo = FALSE, warning = FALSE,cache = TRUE}
data2 <- read.table(stringsAsFactors = FALSE, header = TRUE, sep="/", text =
'Scenario/	Incomplete variable/	Rho/	N; $n_i$/	Missing process
Base/	Continuous/	0.6/	10;1000/	Heckman (BVN)
M(N)AR/Continuous, Binary/ 0 (MAR), 0.3,0.6,0.9 (MNAR)/	10;1000/Heckman (BVN)
Size and cluster number/	Continuous/	0.6/	10;50,   10;100,   10;1000,  50;1000,  100;1000/	Heckman (BVN)
Distribution deviations/	Continuous/	0.6/	10;1000	/Heckman (BVN), Heckman (t-skew), Explicit'
)
options(kableExtra.latex.load_packages = FALSE) 
library(kableExtra)
kbl(data2, booktabs = T, align="l",
    caption = "Data generation scenarios",
    col.names = c("Scenario","Incomplete variable", "$\\rho$", "$N;n_i$","Missing process"),
    row.names=FALSE,escape = FALSE)%>%
    column_spec(c(1), width = "2,5cm")%>%
    column_spec(c(2,4), width = "1,5cm")%>%
    column_spec(c(3), width = "2,9cm")%>%
    column_spec(c(5), width = "2.7cm")%>%
  kable_styling(latex_options = c("striped"))
```
For each dataset, we generated $X_{1i}$ a treatment indicator variable from a Bernoulli distribution with a probability of treatment on each cluster equal to 0.6. Next, we simulated the mean of two continuous covariates from a multivariate normal distribution $\begin{psmallmatrix}\mu_2\\\mu_3\end{psmallmatrix}\sim N\left(\begin{psmallmatrix}0\\0\end{psmallmatrix},\begin{psmallmatrix}0.2 & 0.015\\0.015 & 0.2\end{psmallmatrix}\right)$. We then simulated for each cluster a baseline covariate $X_{2i}\sim N(\mu_2,1)$ and a exclusion restriction variable  $X_{3i} \sim N(\mu_3,0.5)$.

@@

As regards the error terms, they were bivariate normal distributed as:
\[\begin{pmatrix}
\epsilon^O_{i} \\
\epsilon^S_{i}
\end{pmatrix}\sim N\left(\begin{pmatrix}
0 \\
0
\end{pmatrix},\begin{pmatrix}
\sigma^2_i & \rho\sigma_i \\
\rho\sigma_i & 1
\end{pmatrix}\right)
\]
whose $\sigma_i^2$ is a random parameter across clusters distributed as $log(\sigma_i) \sim N(0,0.05)$.
